---
title: "Reproduction Number for Malaysia"
author: "Kamarul Imran Musa"
date: "6/16/2020 updated on `r Sys.Date()`"
output:
  html_document: 
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dpi = 300)
```

# Motivation

To model the time-dependent reproduction number of COVID-19 for Malaysia

# Methods

-   Reproduce methods proposed in EpiEstim package
-   Reference <https://cran.r-project.org/web/packages/EpiEstim/vignettes/demo.html>

Data from

-   WHO from github
-   Crowd-sourcing data from data world

# Prepare environment

```{r}
library(tidyverse)
library(EpiEstim)
library(here)
library(janitor)
library(lubridate)
```

# Data

Steps:

-   Data is downloaded from github
-   save and read again
-   and save locally as covid
-   and read

```{r}
# covid <- read_csv('https://raw.githubusercontent.com/RamiKrispin/coronavirus/master/csv/coronavirus.csv')
# write_csv(covid, 'covid.csv')
covid <- read_csv('coronavirus.csv')
```

## Malaysia data

-   Need to filter country

-   Only select Malaysia

-   Choose data as first **local** case

-   First local cases detected on Feb 4th, 2020 (Day 1)

```{r}
mys <- covid %>% 
  select(date, country, type , cases) %>% 
  filter(country == 'Malaysia', type == 'confirmed') %>% 
  clean_names() %>%
  rename(Time = date, I = cases) %>% 
  filter(Time > '2020-02-3') %>%
  data.frame()
```

```{}
```

### Take data from Day 1 (Feb 4) to Day 100 (May 14)

```{r}
mys <- mys %>% 
  filter(Time < '2020-05-14')
```

# Calculate Time-varying R

## Based on Wuhan data

-   To calculate the Rt, we need to decide on the parameters for serial intervals

-   We set the mean_si = 7.5 and std_si = 3.4. Question: What would be the BEST mean_si and std_si?

-   What would be the earliest time to start and time to end?

-   t_start = need to adjust t_end = need to adjust


```{r}
mys_parametric_si <- estimate_R(mys, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 7.5, 
                                                          std_si = 3.4)))
glimpse(mys_parametric_si)
mys_parametric_si$R
```

### Plot

Will create 

-   plot of the incidence (new confirmed cases). This is the epidemic curve
-   plot the serial interval distribution
-   plot the R (the instantenous R, the time-varying $R_e$)

```{r}
p_I <- plot(mys_parametric_si, "incid") 
p_I + theme_bw() + 
  geom_vline(xintercept = 44, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 58, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 72, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 86, colour = 'red', size = 1.2) +
  geom_text(aes(x = 44, label = '\nMCO Phase 1', y=300),
            colour="blue", angle=90, hjust = 1.2, text=element_text(size=11)) +
  geom_text(aes(x = 58, label = '\nMCO Phase 2', y= 315),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 72, label = '\nMCO Phase 3', y=250),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 86, label = '\nMCO Phase 4', y=250),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  scale_x_continuous(breaks=seq(0, 101, 5)) +
  scale_y_continuous(breaks=seq(0, 350, 25)) +
  ylab('Confirmed new cases') +
  xlab('Number of days from Feb 4, 2020')
```

```{r}
p_SI <- plot(mys_parametric_si, "SI")  
p_SI + theme_bw() +
  scale_x_continuous(breaks=seq(0, 25, 2)) +
  scale_y_continuous(breaks=seq(0, 0.2, 0.025)) +
  xlab('Days')
```

-   the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri <- plot(mys_parametric_si, "R")
p_Ri + theme_bw() + 
  geom_vline(xintercept = 44, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 58, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 72, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 86, colour = 'red', size = 1.2) +
  geom_text(aes(x = 44, label = '\nMCO Phase 1', y=20),
            colour="blue", angle=90, hjust = 1.2, text=element_text(size=11)) +
  geom_text(aes(x = 58, label = '\nMCO Phase 2', y= 20),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 72, label = '\nMCO Phase 3', y=20),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 86, label = '\nMCO Phase 4', y=20),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  scale_x_continuous(breaks=seq(0, 101, 5)) +
  scale_y_continuous(breaks=seq(0, 25, 2)) +
  ylab('Instantaneous R') +
  xlab('Number of days from Feb 4, 2020')
```

The $R_e$ is very high between day 40 to 42. Reasons:

-   One possible explanation is that COVID-19 is transmissible before the onset of symptoms, resulting in much shorter serial intervals than expected, possibly shorter than the incubation period
-   Alternatively, and very likely, there may be non-symptomatic, sub-clinical spreaders of the disease, who are undetected.
-   some cases transmitting the disease very soon after infection, possibly before the onset of symptoms (so-called super-spreaders), and some cases being sub-clinical, and thus undetected, spreading the disease as well, while other cases have a serial interval more consistent with that of MERS or SARS, with a mean around 8 days.

### Calculate $R_e$ with uncertainty

incorporate this uncertainty around the serial interval distribution by allowing specification of a distribution of distributions of serial intervals. So let's

-   retain the mean SI estimated by Li et al of 7.5 days, with an SD of 3.4
-   but let's also allow that mean SI to vary between 2.3 and 8.4
-   Use a truncated normal distribution with an SD of 2.0.
-   We'll also allow the SD or the SD to vary between 0.5 and 4.0.

```{r}
# mys_res_uncertain_si <- estimate_R(mys, method = "uncertain_si", 
#                                   config = make_config(list(mean_si = 7.5, std_si = 3.4, 
#                                                             std_mean_si = 2, 
#                                                             min_mean_si = 2.3, max_mean_si = 8.4, 
#                                                             std_std_si = 2, 
#                                                             min_std_si = 0.5, max_std_si = 4, 
#                                                             n1 = 1000, n2 = 1000)))

# plot_Ri(mys_res_uncertain_si)
```

## Based on Du, Z., et al.

The serial interval of COVID-19 from publicly reported confirmed cases. medRxiv, 2020

- mean = 3.96
- SD = 4.75

```{r}
mys_parametric_si_du <- estimate_R(mys, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 3.96, 
                                                          std_si = 4.75)))
glimpse(mys_parametric_si_du)
```

### Plot

Will create 

-   plot of the incidence (new confirmed cases). This is the epidemic curve
-   plot the serial interval distribution
-   plot the R (the instantenous R, the time-varying $R_e$)

```{r}
p_I_du <- plot(mys_parametric_si_du, "incid") 
p_I_du + theme_bw() + 
  geom_vline(xintercept = 44, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 58, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 72, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 86, colour = 'red', size = 1.2) +
  geom_text(aes(x = 44, label = '\nMCO Phase 1', y=300),
            colour="blue", angle=90, hjust = 1.2, text=element_text(size=11)) +
  geom_text(aes(x = 58, label = '\nMCO Phase 2', y= 315),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 72, label = '\nMCO Phase 3', y=250),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 86, label = '\nMCO Phase 4', y=250),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  scale_x_continuous(breaks=seq(0, 101, 5)) +
  scale_y_continuous(breaks=seq(0, 350, 25)) +
  ylab('Confirmed new cases') +
  xlab('Number of days from Feb 4, 2020')
```

```{r}
p_SI_du <- plot(mys_parametric_si_du, "SI")  
p_SI_du + theme_bw() +
  scale_x_continuous(breaks=seq(0, 25, 2)) +
  scale_y_continuous(breaks=seq(0, 0.8, 0.05)) +
  xlab('Days')
```

-   the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_du <- plot(mys_parametric_si_du, "R")
p_Ri_du + theme_bw() + 
  geom_vline(xintercept = 44, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 58, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 72, colour = 'red', size = 1.2) +
  geom_vline(xintercept = 86, colour = 'red', size = 1.2) +
  geom_text(aes(x = 44, label = '\nMCO Phase 1', y=5),
            colour="blue", angle=90, hjust = 1.2, text=element_text(size=11)) +
  geom_text(aes(x = 58, label = '\nMCO Phase 2', y= 5),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 72, label = '\nMCO Phase 3', y=5),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  geom_text(aes(x = 86, label = '\nMCO Phase 4', y=5),
            colour="blue", angle=90, hjust = 1, text=element_text(size=11)) +
  scale_x_continuous(breaks=seq(0, 101, 5)) +
  scale_y_continuous(breaks=seq(0, 8, 0.5)) +
  ylab('Instantaneous R') +
  xlab('Number of days from Feb 4, 2020')
```

# Analysis from data world

## Wuhan SI

mean SI = 7.5, SD SI = 3.4

```{r}
mys_dw <- mys_dw %>% select(Time, I)
mys_dw_parametric_si <- estimate_R(mys_dw, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 7.5, 
                                                          std_si = 3.4)))
glimpse(mys_dw_parametric_si)
```

### Plot

-   plots the incidence
-   plots the serial interval distribution
-   plot of R

```{r}
p_I_dw <- plot(mys_dw_parametric_si, "incid") 
p_I_dw + theme_bw()
```

```{r}
p_SI_dw <- plot(mys_dw_parametric_si, "SI")  
p_SI_dw + theme_bw()
```

-   the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_dw <- plot(mys_dw_parametric_si, "R")
p_Ri_dw + theme_bw()
```

## Based on Du et al

### Du et al SI

mean SI = 7.5, SD SI = 3.4

```{r}
mys_dw <- mys_dw %>% select(Time, I)
mys_dw_parametric_si_du <- estimate_R(mys_dw, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 3.96, 
                                                          std_si = 4.75)))
glimpse(mys_dw_parametric_si_du)
```

### Plot

-   plots the incidence
-   plots the serial interval distribution
-   plot of R

```{r}
p_I_dw_du <- plot(mys_dw_parametric_si_du, "incid") 
p_I_dw_du + theme_bw()
```

```{r}
p_SI_dw_du <- plot(mys_dw_parametric_si_du, "SI")  
p_SI_dw_du + theme_bw()
```

-   the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_dw_du <- plot(mys_dw_parametric_si_du, "R")
p_Ri_dw_du + theme_bw()
```

# Values of R at MCO 1, 2 and 3

-   First case local Malaysia on Feb 4 (<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7293423/>)
-   MCO 1 on 18 March - 31 March
-   MCO 2 on 1 April - 14 April
-   MCO 3 on 15 April - 28 April
-   MCO 4 on 29 April - 12 May

Calculate difference in days

```{r}
first_mco1 <- interval(as_date('2020-02-04'), as_date('2020-03-17'))
first_mco1 <- first_mco1 / ddays(1)

mco1_mco2 <- interval(as_date('2020-03-17'), as_date('2020-03-31'))
mco1_mco2 <- mco1_mco2 / ddays(1)

mco2_mco3 <- interval(as_date('2020-03-31'), as_date('2020-04-14'))
mco2_mco3 <- mco2_mco3 / ddays(1)

mco3_mco4 <- interval(as_date('2020-04-14'), as_date('2020-04-28'))
mco3_mco4 <- mco3_mco4 / ddays(1)

mco4_mco5 <- interval(as_date('2020-04-28'), as_date('2020-05-12'))
mco4_mco5 <- mco4_mco5 / ddays(1)

first_mco1 ; mco1_mco2 ; mco2_mco3 ; mco3_mco4 ; mco4_mco5
```

In the table \@ref(tab:si), we see the estimated R value from the first day until .

```{r si}
mys_dw_parametric_si$R
```

In the table \@ref(tab:mco), we show for each MCO phase, the difference in the mean R value and the percentage of difference in means between

1.  day 10 and day 5
2.  day 5 and day 1 .

```{r mco}
MCO <- mys_dw_parametric_si$R %>% 
  filter(t_start %in% c(2, 7, 12, 55, 60, 65, 68, 73, 78, 80, 85, 88))
MCO
MCO_type <- c(rep('MCO1',3), 
              rep('MCO2',3),
              rep('MCO3',3),
              rep('MCO4',3)) 
data_mco <- data.frame(day =  c('D1', 'D5' , 'D10'), MCO_type)
data_mco <- bind_cols(data_mco, MCO) %>% rename(meanR = "Mean(R)")
data_mco
```

Calculate the difference

```{r}
data_mco_R <- data_mco %>%
    group_by(MCO_type) %>%
    mutate(mean_diff = meanR - lag(meanR),
           mean_diff_percent = (mean_diff / (meanR - mean_diff))*100,
           day = as_factor(day)) %>% 
  select(day, MCO_type, meanR, mean_diff, mean_diff_percent, t_start, t_end, 
         quantile25 = `Quantile.0.025(R)`, quantile975 = `Quantile.0.975(R)`)
data_mco_R
```

plot

```{r}
data_mco_R %>% ggplot(aes(x = day, y = meanR, group = MCO_type)) + 
  geom_errorbar(aes(ymin = quantile975, ymax = quantile25), width = .25) +
  geom_line() +
  geom_point() +
  facet_wrap(. ~ MCO_type, ncol = 2, scales = 'free_y') +
  ylab('mean reproduction number and quantiles') +
  ggtitle('The reproduction number at day 1 (D1), day 5 (D5) and day 10 (D10) \nand the 25% quantiles to 95% quantiles intervals')
```

```{r}
data_mco_R %>% ggplot(aes(x = day, y = meanR, group = MCO_type)) + 
  geom_errorbar(aes(ymin = quantile975, ymax = quantile25), width = .25) +
  geom_line() +
  geom_point() +
  facet_grid(. ~ MCO_type) +
  ylab('mean reproduction number and quantiles') +
  ggtitle('The reproduction number at day 1 (D1), day 5 (D5) and day 10 (D10) \nand the 25% quantiles to 95% quantiles intervals')
```

```{r}
sessionInfo()
```
