---
title: "Reproduction Number for Malaysia"
author: "Kamarul Imran Musa"
date: "6/16/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Motivation

To model the time-dependent reproduction number of COVID-19 for Malaysia

# Methods

- Reproduce methods proposed in EpiEstim package
- Reference  https://cran.r-project.org/web/packages/EpiEstim/vignettes/demo.html

Data from

- WHO from github
- Crowd-sourcing data from data world

# Prepare environment

```{r}
library(tidyverse)
library(EpiEstim)
library(here)
library(janitor)
library(lubridate)
```


# Data

## From github, save and read again

and save locally as covid
and read

```{r}
# covid <- read_csv('https://raw.githubusercontent.com/RamiKrispin/coronavirus/master/csv/coronavirus.csv')
# write_csv(covid, 'covid.csv')
covid <- read_csv('covid.csv')
```

## Malaysia data

Malaysia data from github

mys starts from first local case (Feb 4th)

```{r}
mys <- covid %>% 
  select(date, country, type , cases) %>% 
  filter(country == 'Malaysia', type == 'confirmed') %>% clean_names() %>%
  rename(Time = date, I = cases) %>% filter(Time > '2020-02-3') %>%
  data.frame()
```

Malaysia data from dataworld

```{r}
library("httr")
library("readxl")
GET("https://query.data.world/s/dl2knkmq7y2erjews5hnsc6vjbr3pc", 
    write_disk(tf <- tempfile(fileext = ".xlsx")))
mys_dw <- read_excel(tf, sheet = 'd(Base)') %>% clean_names() %>%
  select(Time = date, mco_phase, cum_pos = positive, I = d_positive) %>%
  mutate_if(is.POSIXt, as.Date) %>% 
  filter(Time < '2020-06-07') %>%
  data.frame()
mys_dw_state <- read_excel(tf, sheet = 'State') %>% clean_names() %>%
  select(Time = date, perlis:total, east_malaysia, west_malaysia) %>%
  mutate_if(is.POSIXt, as.Date) %>% filter(Time > '2020-02-3') %>%
  data.frame()
```

# Calculate R

## Based on Wuhan

Setting the mean_si = 7.5 and std_si = 3.4. What would be the mean_si and std_si?

The problem is R is estimated too early.

What would be the earliest time to start and time to end?

t_start = need to adjust
t_end = need to adjust

For Malaysia data John Hopkins 

```{r}
mys_parametric_si <- estimate_R(mys, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 7.5, 
                                                          std_si = 3.4)))
glimpse(mys_parametric_si)
```
### Plot

- plots the incidence
- plots the serial interval distribution
- plot of R

```{r}
p_I <- plot(mys_parametric_si, "incid") 
p_I + theme_bw()
```

```{r}
p_SI <- plot(mys_parametric_si, "SI")  
p_SI + theme_bw()
```
- the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri <- plot(mys_parametric_si, "R")
p_Ri + theme_bw()
```

The $R_t$ is very high nearly day 50. Reasons:

- One possible explanation is that COVID-19 is transmissible before the onset of symptoms, resulting in much shorter serial intervals than expected, possibly shorter than the incubation period
- Alternatively, and very likely, there may be non-symptomatic, sub-clinical spreaders of the disease, who are undetected.
- some cases transmitting the disease very soon after infection, possibly before the onset of symptoms (so-called super-spreaders), and some cases being sub-clinical, and thus undetected, spreading the disease as well, while other cases have a serial interval more consistent with that of MERS or SARS, with a mean around 8 days.

### Calculate R with uncertainty

incorporate this uncertainty around the serial interval distribution by allowing specification of a distribution of distributions of serial intervals. So let’s 

- retain the mean SI estimated by Li et al of 7.5 days, with an SD of 3.4
- but let’s also allow that mean SI to vary between 2.3 and 8.4 
- Use a truncated normal distribution with an SD of 2.0.
- We’ll also allow the SD or the SD to vary between 0.5 and 4.0.

```{r}
# mys_res_uncertain_si <- estimate_R(mys, method = "uncertain_si", 
#                                   config = make_config(list(mean_si = 7.5, std_si = 3.4, 
#                                                             std_mean_si = 2, 
#                                                             min_mean_si = 2.3, max_mean_si = 8.4, 
#                                                             std_std_si = 2, 
#                                                             min_std_si = 0.5, max_std_si = 4, 
#                                                             n1 = 1000, n2 = 1000)))

# plot_Ri(mys_res_uncertain_si)
```


## Based on Du, Z., et al.

The serial interval of COVID-19 from publicly reported confirmed cases. medRxiv, 2020

mean = 3.96, SD = 4.75

```{r}
mys_parametric_si_du <- estimate_R(mys, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 3.96, 
                                                          std_si = 4.75)))
glimpse(mys_parametric_si_du)
```
### Plot

- plots the incidence
- plots the serial interval distribution
- plot of R

```{r}
p_I_du <- plot(mys_parametric_si_du, "incid") 
p_I_du + theme_bw()
```

```{r}
p_SI_du <- plot(mys_parametric_si_du, "SI")  
p_SI_du + theme_bw()
```

- the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_du <- plot(mys_parametric_si_du, "R")
p_Ri_du + theme_bw()
```



# Analysis from data world

## Wuhan SI

mean SI = 7.5, SD SI = 3.4 

```{r}
mys_dw <- mys_dw %>% select(Time, I)
mys_dw_parametric_si <- estimate_R(mys_dw, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 7.5, 
                                                          std_si = 3.4)))
glimpse(mys_dw_parametric_si)
```

### Plot

- plots the incidence
- plots the serial interval distribution
- plot of R

```{r}
p_I_dw <- plot(mys_dw_parametric_si, "incid") 
p_I_dw + theme_bw()
```


```{r}
p_SI_dw <- plot(mys_dw_parametric_si, "SI")  
p_SI_dw + theme_bw()
```

- the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_dw <- plot(mys_dw_parametric_si, "R")
p_Ri_dw + theme_bw()
```


## Based on Du et al

### Du et al SI

mean SI = 7.5, SD SI = 3.4 

```{r}
mys_dw <- mys_dw %>% select(Time, I)
mys_dw_parametric_si_du <- estimate_R(mys_dw, 
                                method = "parametric_si",
                                config = make_config(list(mean_si = 3.96, 
                                                          std_si = 4.75)))
glimpse(mys_dw_parametric_si_du)
```

### Plot

- plots the incidence
- plots the serial interval distribution
- plot of R

```{r}
p_I_dw_du <- plot(mys_dw_parametric_si_du, "incid") 
p_I_dw_du + theme_bw()
```


```{r}
p_SI_dw_du <- plot(mys_dw_parametric_si_du, "SI")  
p_SI_dw_du + theme_bw()
```

- the 7-day sliding window estimates of instantaneous $R_e$

```{r}
p_Ri_dw_du <- plot(mys_dw_parametric_si_du, "R")
p_Ri_dw_du + theme_bw()
```
# Values of R at MCO 1, 2 and 3

- First case local Malaysia on Feb 4 (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7293423/)
- MCO 1 on 18 March - 31 March
- MCO 2 on 1 April - 14 April
- MCO 3 on 15 April - 28 April
- MCO 4 on 29 April - 12 May

Calculate difference in days

```{r}
first_mco1 <- interval(as_date('2020-02-04'), as_date('2020-03-17'))
first_mco1 <- first_mco1 / ddays(1)

mco1_mco2 <- interval(as_date('2020-03-18'), as_date('2020-03-31'))
mco1_mco2 <- mco1_mco2 / ddays(1)

mco2_mco3 <- interval(as_date('2020-04-01'), as_date('2020-04-14'))
mco2_mco3 <- mco2_mco3 / ddays(1)

mco3_mco4 <- interval(as_date('2020-04-15'), as_date('2020-04-28'))
mco3_mco4 <- mco3_mco4 / ddays(1)

mco4_mco5 <- interval(as_date('2020-04-29'), as_date('2020-05-12'))
mco4_mco5 <- mco4_mco5 / ddays(1)

first_mco1 ; mco1_mco2 ; mco2_mco3 ; mco3_mco4 ; mco4_mco5
```


```{r}
mys_dw_parametric_si$R
```


```{r}
sessionInfo()
```

